---
title: "Comparison of duplicated genes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dev = 'svg')
```

```{r loadPackages, include=FALSE}
library(tidyverse)
library(ggplot2)
library(plotly)
library("scales")

# source("R/calcRanks.R")
source("R/orthoUtils.R")
# source("R/divUtils.R")

```

## Aim of this analysis

investigate the expression similarity of duplicated genes

```{r loadRankOrthoTables}

orthoRankTables <-
  dir("data/ranks", pattern="...._orthoRanks.RDS",full.names=T)  %>%
  set_names(sub(".*/(....)_orthoRanks.RDS","\\1", .)) %>% 
  map(readRDS) %>% 
  map( function( orthoTable ){
    # anonymize the spc1 and spc2 columns in the table
    names(orthoTable)[1:2] <- c("SPC1", "SPC2")
    
    return(orthoTable)
  })

# calculate size of groups
grpSize <-
  lapply(orthoRankTables, function( orthoTable ){
    orthoTbl2Array(orthoTable) %>%
      as_data_frame() %>% 
      mutate( N1 = map_int(SPC1,length),
              N2 = map_int(SPC2,length))
  })


```

## Rank distribution and ortholog duplication level

The following plots show the distribution of ranks (i.e. -log10( Pval ) ) for each pair of species and for ortholog duplication level. "N1" refers to the number of ortholog copies in the first species and N2 in the second. (e.g. a 1:2 ortholog in At:Os would be AtOs N1=1 N2=2)


```{r plotOrthoRanks, fig.height=9,fig.width=15}
x <- 
  # add grpSize to orthoRankTables and concatenate into one table
  map2_df(grpSize, orthoRankTables, .id="spcPair",
        ~ .x %>% 
          unnest(SPC1) %>% 
          inner_join(.y, by = "SPC1")) %>% 
  # gather rnksT and rnks to one column
  gather( rnks, rnksT, key="rnkT", value="rnk") %>% 
  # remove NA's
  filter( !is.na(rnk), N1 < 6, N2 < 6) %>% 
  # log transform ranks
  mutate( logRnk = -log10(1.0001-rnk) ) 

x %>%  
  mutate( N1=as.factor(N1), N2=as.factor(N2)) %>% 
  ggplot( aes( x=N1, y=logRnk) ) +
  facet_grid(N2 ~ as.factor(spcPair)) +
  geom_violin() +
  stat_summary(fun.y=median, geom="point", size=2, shape=20)

```
The following heatmap of median ranks which may give a better overview:

```{r orthoMedRnkHeatmap}
x %>% 
  mutate( N1=as.factor(N1), N2=as.factor(N2)) %>% 
  group_by(spcPair,N1,N2) %>% 
  summarise( medRnk = median(logRnk)) %>% 
  ggplot() +
  geom_tile( aes(x=N1,y=N2,fill=medRnk) ) +
  facet_wrap( ~ as.factor(spcPair), nrow=2 ) +
  scale_fill_gradientn( colors = RColorBrewer::brewer.pal(5,"Spectral"),
                       values = scales::rescale(4:0),
                       limits = c(0,2)) +
  coord_equal() +
  theme_bw()
  

```

The main observation is that 1:1 orthologs clearly tend to have higher ranks than orthologs with any number of duplicates and that the ranks get lower when the number of orthologs increases. (Note that there is increased uncertainty in the median of the larger groups as they are more rare). An interresting pattern is that orthologs with either one or two copies in Gm seem to have the approximately same ranks. A possible explanation could be the recent duplication in Gm and that a high portion of the genes that prefer occurr in single copies haven't had time to lose its copy.

```{r OrthoRankTrend}
reverselog_trans <- function(base = exp(1)) {
    trans <- function(x) -log(x, base)
    inv <- function(x) base^(-x)
    trans_new(paste0("reverselog-", format(base)), trans, inv, 
              log_breaks(base = base), 
              domain = c(1e-100, Inf))
}

# take only the 1:1-4 and plot in a single figure
x %>% 
  # Swap the order of spc so that the spc with a single ortholog comes first (e.g. AtOs 1:2 = At1:Os2, OsAt 1:2 = Os1:At2)
  mutate( spcPair=ifelse(N2<N1,sub("(..)(..)","\\2\\1",spcPair), spcPair) ) %>%
  mutate( Pval = 1.0001-rnk ) %>% 
  group_by(spcPair,N1,N2,rnkT) %>% 
  summarise( medPval = median(Pval)) %>% 
  # summarise( medRnk = median(logRnk)) %>% 
  filter( (N1==1 & N2<5) | (N2==1 & N1<5) ) %>%
  mutate( orthoDupLevel = paste0("1:",pmax(N1,N2))) %>%
  ggplot() +
  # geom_line( aes(x=pmax(N1,N2),y=medPval,color=spcPair,linetype=rnkT) ) +
  geom_point( aes(x=as.factor(paste0("1:",pmax(N1,N2))),y=medPval) ) +
  theme_bw() +
  ylab("Median P-value") +
  xlab("Ortholog count") +
  scale_y_continuous(breaks=c(1,0.5,0.1,0.05,0.01),trans=reverselog_trans(10)) 
  
  #scale_x_continuous(breaks=1:4, labels=paste0("1:",1:4)) -> g

# ggplotly(g)

```

## 1:2 orthologs

If we look at 1:2 orthologs we can plot the score between the single ortholog and the two duplicate orthologs. The following plots show the -log10(P) transformed rank for each of the two alternative orthologs plotted against each other. If both of the duplicate genes have retained the same function we expect them to be similar, while if one of the duplicates have taken a different function (i.e. neo- or sub-functionalized) they may have diverged expressed patterns. Since the score for each of the duplicates also depends on whether the single ortholog has diverged it is however expected the scores correlated to a certain degree. There is one plot for each species-pair and the order of the species indicates which species that contain the duplicate, e.g. AtGm contains the At1:Gm2 orthologs.

```{r}
library(ggpmisc) # stat_poly_eq

# 1:2
x %>%
  filter( N1==1, N2==2, rnkT=="rnks") %>% 
  # annotate spcPar e.g. "AtGm" becomes "At1:Gm2"
  mutate( spcPair = sub("(..)(..)","\\11:\\22",spcPair)) %>% 
  # for each single ortholog (geneID in spc1)
  group_by(SPC1,spcPair) %>%
  # separate the ranks for first and second (i.e. last) duplicate
  summarise( rnk1=first(logRnk),rnk2=last(logRnk))  %>%
  ggplot(aes(x=rnk1,y=rnk2)) +
  ylab("-log10(P) of duplicate 2") +
  xlab("-log10(P) of duplicate 1") +
  geom_point() +
  geom_density2d() +
  coord_equal() +
  stat_poly_eq( formula = x~y,
                aes(label = ..rr.label..), fontface="bold",color="red",
                parse = TRUE) +
  facet_wrap( ~ as.factor(spcPair), nrow=3 )

# 2:1
x %>%
  filter( N1==2, N2==1, rnkT=="rnksT") %>% 
  mutate( spcPair = factor(sub("(..)(..)","\\21:\\12",spcPair),
                           ordered = T,
                           levels = sub("(..)(..)","\\21:\\12",unique(spcPair)))) %>% 
  group_by(SPC2,spcPair) %>%
  summarise( rnk1=first(logRnk),rnk2=last(logRnk)) %>% 
  ggplot(aes(x=rnk1,y=rnk2)) +
  ylab("-log10(P) of duplicate 2") +
  xlab("-log10(P) of duplicate 1") +
  geom_point() +
  geom_density2d() +
  coord_equal() +
  stat_poly_eq( formula = x~y,
                aes(label = ..rr.label..), fontface="bold",color="red",
                parse = TRUE) + 
  facet_wrap( ~ spcPair, nrow=3 )


```

## At1:Gm2 

The set of genes that have a 1:1 ortholog relationship between At and Os are more likely to be essential and therefore have a conserved expression pattern. Given the recent WGD in Gm there are many of these genes that have duplicate orthologs in Gm (i.e. At1:Gm2). The plots below compare the ranks distribution of two groups of At1:Gm2 orthologs, those that are At1:Os1 (singleton) and those that are not. The singletons are expected to be under selective pressure to retain the expression pattern.

There seems to be a lower correlation between Gm duplicates among the singletons, perhaps an indication of a selective pressure to conserve only one of the duplicates (i.e. one duplicate is on the way to non-functionalizaition).

However, there is a also significant difference in the mean ranks (see wilcoxon rank-sum test below) of the two duplicates in the two groups, i.e. the singleton group tend to have higher ranks in general. This might indicate that most of the expression divergence occured in the At lineage or before the WGD in the Gm lineage.


```{r testGmDuplicateWithOsSingleton}
At1Os1 <-
  filter(x, spcPair=="AtOs", N1==1, N2==1, rnkT=="rnksT")

At1Gm2 <-
  filter(x, spcPair=="AtGm", N1==1, N2==2) %>% 
  group_by(SPC1, rnkT) %>%
  summarise( meanRnk=mean(logRnk), 
             rnk1=first(logRnk),
             rnk2=last(logRnk) ) %>% 
  mutate( inAt1Os1 = SPC1 %in% At1Os1$SPC1) %>% 
  mutate( isAt1Os1 = as.factor(ifelse(inAt1Os1, "At:Os singleton","not At:Os singleton")) )


At1Gm2 %>% 
  mutate( ORSdir = as.factor(c(rnks="At -> Gm", rnksT="Gm -> At")[rnkT]) ) %>% 
  ggplot(aes(x=rnk1,y=rnk2)) +
  ggtitle("At1:Gm2 orthologs sorted by singleton state in Os") +
  ylab("-log10(P) of duplicate 2") +
  xlab("-log10(P) of duplicate 1") +
  geom_point() +
  geom_density2d() +
  coord_equal() +
  stat_poly_eq( formula = x~y,
                aes(label = ..rr.label..), fontface="bold",color="red",
                parse = TRUE) +
  facet_grid( ORSdir ~ isAt1Os1 )



```

```{r}

W <- wilcox.test(x = At1Gm2$meanRnk[At1Gm2$inAt1Os1],
            y = At1Gm2$meanRnk[!At1Gm2$inAt1Os1],
            alternative = "greater" )
At1Gm2 %>% 
  ggplot( aes(x=meanRnk, color=isAt1Os1)) + 
  ggtitle( paste("Wilkox rank-sum test P =",signif(W$p.value,2))) +
  xlab("Mean -log10(P) of At1:Gm2 orthologs") +
  geom_density()
```


If we compare the ranks of the At1:Os1 orthologs with the corresponding mean ranks of the At1:Gm2 orthologs we see that they correlate well but that At1:Gm2 rarely have higher rank than At1:Os1. As At:Os is more distantly related than At:Gm this is not expected. A boring explanation could be that the reduced ranks in At:Gm is mainly a result of sample bias in Gm.

```{r}
At1Os1 %>% 
  transmute(SPC1=SPC1, logRnkAtOs=logRnk) %>% 
  inner_join(At1Gm2, by="SPC1") %>% 
  ggplot( aes(x=logRnkAtOs, y=meanRnk)) +
  xlab("rank At1Os1") +
  ylab("mean rank At1Gm2") +
  geom_point() +
  coord_equal() +
  stat_poly_eq( formula = x~y,
                aes(label = ..rr.label..), fontface="bold",color="red",
                parse = TRUE) +  
  geom_density2d()
```



```{r eval=F}

grpSize$AtGm %>% filter(N2==2, N1==1) %>% select(SPC1, SPC2) %>% .[1,] %>% as.list()



treeID <- geneID2treeID$At["AT5G08400"]

p4d <- treeDataToPhylo4d(fixTreeData(treeData[[treeID]]))
p4d <- p4d[tipData(p4d)$taxonID %in% spc2taxid]
tr <- getPhylo(p4d)

library(ape)
plot.phylo( nameTree(phylo =  tr))
plot.phylo( tr)
nodelabels(nodeData(p4d)$taxonName,frame = "none",adj=c(0,0.5),cex=0.6)
markDupNodes(p4d)


```



