---
title: "Effect of number of samples included"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dev = 'svg')
```

```{r loadPackages, include=FALSE}
library(tidyverse)
source("R/divUtils.R") # unnest1
```

### aim of the analysis

Illustrate how the number of samples affect the ortholog rank score (ORS).

### Method

To test the effect of the number of samples we picked subsets of various sizes out of the 1363 _At_ samples available and calculated ORS (MI+CLR) against the full set of _Os_ samples (454 samples). Only 1:1 orthologs were included (~5k genes) to save computation time.

Samples were randomized in two different ways, either by shuffling the order of the individual samples or by shuffling the order of the studies (i.e. sets of samples from the same experiment). Note that keeping the samples from each study together is closer to the realistic situation where all available data are used. Using the randomized sample order, the first 12,25,50,100,200,400 and 800 samples were used to calculate ORS. This was repeated 10 and 5 times, for study and individual sample randomization respectively.

### Result

```{r loadData}

rnks_allSamples <- readRDS("data/subsets/nSamples/rnks_allSamples.RDS")

medRnkAllSamples <- 
  data_frame(rnks = list(rnks_allSamples)) %>% 
  unnest1( .id = "rnkDir" ) %>% 
  mutate( medRnk=map_dbl(rnks, median, na.rm=T))

rnks <- 
  data_frame( filename = dir("data/subsets/nSamples",
                             pattern="rnks_(.*)_(.*)_(.*).RDS",
                             full.names = T)) %>%
  mutate( rnks = map(filename,readRDS),
          rndType = sub("rnks_(.*)_(.*)_(.*).RDS","\\1",basename(filename)),
          idx = as.integer(sub("rnks_(.*)_(.*)_(.*).RDS","\\2",basename(filename))),
          n = as.integer(sub("rnks_(.*)_(.*)_(.*).RDS","\\3",basename(filename))),
          filename = NULL) %>% 
  unnest1( .id = "rnkDir" ) %>% 
  mutate( medRnk=map_dbl(rnks, median, na.rm=T))

# fill in the last point (i.e copy)
rnks <- 
  rnks %>% 
  dplyr::select(idx, rndType, rnkDir) %>%
  unique() %>% 
  mutate( n=1363 ) %>% 
  left_join(medRnkAllSamples, by = "rnkDir") %>% 
  bind_rows(rnks)

```

When randomizing by study, the median ORS always increase when the number of samples increases. In contrast, when randomizing by individaul samples, the ORS seem to reach the maximum and stay level after just one tenth of the samples have been included. Note that, although seldom, there are points were the additional samples results in a lower median ORS.

__Figure text:__ Median ORS between _At_ and _Os_ using various number of _At_ samples. Each of the thin lines represent one sample order randomizition, where each point represents the same samples as the previous point plus addition samples. The thick lines show the mean of the randomizations.

```{r plot_nSamples}
rnks %>% 
  filter(rnkDir=="AtOs") %>% 
  mutate(rndType=factor(rndType,levels = c("rndSamples","rndStudies"), 
                        labels=c("Individual samples","Studies"))) %>% 
  # mutate( logMedRnk = -log10(1-medRnk), medRnkP = 1-medRnk) %>% 
  ggplot( aes(n, medRnk, group = interaction(idx, rndType))) +
  geom_line( size = 0.2, alpha=0.5,
             aes( group = interaction(idx, rndType),
                  color = rndType)) +
  geom_point( size = 0.2, alpha=0.5,
             aes( group = interaction(idx, rndType),
                  color = rndType)) +
  stat_summary( fun.y = mean, geom = "line", size = 0.7,
                aes(color = rndType, group = rndType)) +
  stat_summary( fun.y = mean, geom = "point", size= 2,
                aes(color = rndType, group = rndType)) +
  scale_colour_brewer(palette = "Set1") +
  scale_x_log10( breaks=c(10,20,50,100,200,500,1000),
                 minor_breaks=c((1:10)*10^1,(1:10)*10^2)) +
  ylim(c(NA,1)) +
  xlab("Number of At samples included") +
  ylab("Median ortholog rank score") +
  labs(color = "Randomized by:") +
  theme_linedraw() +
  theme(legend.position = "bottom")
```

The figure below shows the same as above but with ORS for both directions (At -> Os and Os -> At). 

```{r plot_nSamples_withRnkDir}
rnks %>% 
  mutate(rndType=factor(rndType,levels = c("rndSamples","rndStudies"), 
                        labels=c("Individual samples","Studies"))) %>% 
  # mutate( logMedRnk = -log10(1-medRnk), medRnkP = 1-medRnk) %>% 
  ggplot( aes(n, medRnk, group = interaction(idx, rndType, rnkDir))) +
  geom_line( size = 0.2, alpha=0.5,
             aes( group = interaction(idx, rndType, rnkDir),
                  color = rndType, linetype = rnkDir)) +
  geom_point( size = 0.2, alpha=0.5,
             aes( group = interaction(idx, rndType),
                  color = rndType)) +
  stat_summary( fun.y = mean, geom = "line", size = 0.7,
                aes(color = rndType, linetype = rnkDir, group = interaction(rndType, rnkDir))) +
  stat_summary( fun.y = mean, geom = "point", size= 2,
                aes(color = rndType, group = interaction(rndType, rnkDir))) +
  scale_colour_brewer(palette = "Set1") +
  scale_x_log10( breaks=c(10,20,50,100,200,500,1000),
                 minor_breaks=c((1:10)*10^1,(1:10)*10^2)) +
  ylim(c(NA,1)) +
  xlab("Number of At samples included") +
  ylab("Median ortholog rank score") +
  labs(color = "Randomized by:") +
  theme_linedraw() +
  theme(legend.position = "bottom")

```

Alternative plot with -log10(1-ORS) transformed values:

```{r plot_nSamples_withRnkDirlog}
rnks %>% 
  mutate(rndType=factor(rndType,levels = c("rndSamples","rndStudies"), 
                        labels=c("Individual samples","Studies"))) %>% 
  mutate( logMedRnk = -log10(1-medRnk), medRnkP = 1-medRnk) %>% 
  ggplot( aes(n, logMedRnk, group = interaction(idx, rndType, rnkDir))) +
  geom_line( size = 0.2, alpha=0.5,
             aes( group = interaction(idx, rndType, rnkDir),
                  color = rndType, linetype = rnkDir)) +
  geom_point( size = 0.2, alpha=0.5,
             aes( group = interaction(idx, rndType),
                  color = rndType)) +
  stat_summary( fun.y = mean, geom = "line", size = 0.7,
                aes(color = rndType, linetype = rnkDir, group = interaction(rndType, rnkDir))) +
  stat_summary( fun.y = mean, geom = "point", size= 2,
                aes(color = rndType, group = interaction(rndType, rnkDir))) +
  scale_colour_brewer(palette = "Set1") +
  scale_x_log10( breaks=c(10,20,50,100,200,500,1000),
                 minor_breaks=c((1:10)*10^1,(1:10)*10^2)) +
  xlab("Number of At samples included") +
  ylab("Median of -log10(1-ORS)") +
  labs(color = "Randomized by:") +
  theme_linedraw() +
  theme(legend.position = "bottom")

```

### Discussion

The results show that including more samples is almost always better. Interrrestingly, it also shows that similar performance can be achieved by using only a tenth of the samples when they are drawn randomly from the full set of studies. This indicates that diversity rather than just numbers is important. (As most studies contain replicates and control for certain conditions they will limit the number of possible regulatory states.)

Other co-expression studies have used methods were samples are filtered and/or weighted to improve the results. It would be interresting to test whether such methods could give higher ORS than by simply inlcuding all samples.

I would like to point out that in this test the ORS is limited by the Os samples and by the actual biological differnces between Os and At.