---
title: "Within species comparison"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loadPackages, include=FALSE}
library(plyr)
library(tidyverse)

source("R/divUtils.R") # unnest1
```

## Aim of the analysis

Compare sub-sets of samples from the same species to quantify the effect of the sample selection alone.

## Random sample splitting

For each species the samples are split into two sub-sets by randomly selecting half of the studies. This ensures that for each of the sub-set pairs, the two halves do not share any samples from the same study (such as biological replicates). Note that the number of samples in each sub-set will not be the same. Splitting was repeated 10 times to generate 10 sub-set pairs (later refered to as reps or sample sub-sets). The plot below gives an overview of the sample were split into sub-set pairs:

```{r sampleSelection, echo=F}

# load subsets
ss <- 
  dir("data/subsets/withinSpecies",pattern = "sampleSubset.RDS",
      recursive = T,full.names = T) %>% 
  set_names(sub(".*/(..)/sampleSubset.RDS","\\1",.)) %>% 
  map( readRDS )

for( spc in names(ss)){
  m <- ss[[spc]]$subsetMat
  studySize <- table(ss[[spc]]$studyID)
  x1 <- cumsum(studySize)
  x0 <- c(0,x1[-length(x1)])
  plot(NULL,xlim=c(0,sum(studySize)),ylim=c(1,ncol(m)),
       xlab="sample index",ylab="replicate index", main=spc)
  for(i in 1:ncol(m)){
    segments(x0 = x0[m[ ,i]], x1 = x1[m[ ,i]], y0 = i, y1 = i)
  }
}

```


## Gene selection

To reduce the computational cost, only a subset (n=4000) of the genes were included. Genes with 1:1 orthologs to several species are prioritized and if that is not enough (as is the case for Gm and Zm) then also include 2:1 orthologs.

## Reference "ortholog" selection

When comparing within species there are no orthologs and any gene can be used as a reference "ortholog". To make the comparison more similar to that of the between species we use the 1:1 orthologs to each of the other species as reference orthologs. Ranks are calculated once for each set of 1:1 orthologs against the other species. In later plots "refOrthoSpc" refer to the different sets of orthologs.

## results

The plot below shows the distribution of the resulting ranks for all within-species subsets (10reps x 4reforthos x 2dir x 4000genes) together with the ranks from the between species tests. Median is marked with diamond. Ranks are shown in -log10(P) scale (i.e. a value >1 corresponds P<0.1 and a value >2 corresponds P<0.01 etc.). As expected the within-species ranks are higher than between species.


```{r loadRanks}
#
# load within species ranks
#
# This is a table with 10reps x 5spcs rows where the rnks column contain a
# nested list of 4refspcs x 2dir x 4000genes
withinRnks <- 
  data_frame( filename = dir("data/subsets/withinSpecies",
                      recursive = T,pattern="rnks",full.names = T)) %>% 
  mutate( spc = sub(".*/(..)/rnks_[0-9]+.RDS","\\1",filename),
          repNr = sub(".*/../rnks_([0-9]+).RDS","\\1",filename),
          rnks = map(filename, readRDS))

# flatten the nested list
withinRnksUnnested <- 
  withinRnks %>%
  mutate(spc1=spc, spc2=spc, typ="Within species",
         spc=NULL, filename=NULL) %>% 
  mutate(rnks = map(rnks,unlist)) %>% 
  unnest()

# load Full CCS ranks (including 1:N and N:N)
fullRnks <-
  data_frame( filename=dir("data/ranksMR",pattern="...._ranks.RDS",full.names = T)) %>% 
  mutate(spc1 = sub(".*/(..)(..)_ranks.RDS","\\1",filename),
         spc2 = sub(".*/(..)(..)_ranks.RDS","\\2",filename),
         rnks = map(filename,readRDS),
         filename = NULL)

# Filter only 1:1 and divide into transposed and not
fullRnksT <-
  fullRnks %>% 
  # filter 1:1 and select rnksT as rnks
  mutate( rnks = map(rnks, ~ .x %>% filter(otype=="1:1") %>% transmute(rnks = rnksT)) ) %>% 
  mutate( spc1_=spc1, spc1=spc2, spc2=spc1_) %>% # swap spc1 and spc2
  select(-spc1_) # remove temporary column

fullRnks <- 
  fullRnks %>% 
  # filter 1:1 and select rnks
  mutate( rnks = map(rnks, ~ .x %>% filter(otype=="1:1") %>% transmute(rnks = rnks)) ) %>% 
  bind_rows(fullRnksT) # combine with Transposed ranks

```


```{r fullRankscompare, fig.height=4, fig.width=9, dev='svg', echo=F}
fullRnks %>% 
  unnest() %>% 
  mutate( typ="Between species" ) %>% 
  bind_rows( withinRnksUnnested ) %>% 
  filter( !is.na(rnks) ) %>%  # remove NA values
  mutate( logRnks = -log10(1.0001-rnks)) %>% 
  ggplot( aes( x=spc2, y=logRnks, colour=typ) ) +
  facet_wrap(~as.factor(spc1), nrow=1) +
  geom_violin() +
  stat_summary(fun.y=median, geom="point", size=2, shape=23)


```


```{r calcMedRnkMat}

spcs <- unique(fullRnks$spc1)
medRnkMat <- sapply(spcs,function(spc1){
  sapply(spcs,function(spc2){
    if(spc1==spc2){
      # for each replicate
      mean( sapply(withinRnks$rnks[withinRnks$spc==spc1],function(x){
        # for each ref spc
        sapply(x,function(y){
          # median of mean of direction
          median(y[[1]]+y[[2]])/2
        })
      }))
    } else {
      idx1 <- which(fullRnks$spc1==spc1 & fullRnks$spc2==spc2)
      idx2 <- which(fullRnks$spc1==spc2 & fullRnks$spc2==spc1)
      median(fullRnks$rnks[[idx1]]$rnks+fullRnks$rnks[[idx2]]$rnks,na.rm = T)/2
    }
  })
})

logMedRnkMat <- -log10(1.0001-medRnkMat)
range(logMedRnkMat)
normMedRnkMat <- logMedRnkMat/sqrt(diag(logMedRnkMat)) %o% sqrt(diag(logMedRnkMat))
```

```{r}
myPalette <- colorRampPalette(rev(RColorBrewer::brewer.pal(11, "Spectral")), space="Lab")

logMedRnkMat[c("At","Sl","Gm","Os","Zm"),c("At","Sl","Gm","Os","Zm")] %>% 
  reshape2::melt() %>% 
  ggplot(aes(x = Var2, y = Var1, fill = value)) +
  geom_tile() +
  scale_fill_gradientn(colours = myPalette(100),limits=c(0, 3)) +
  coord_equal() +
  theme_bw()
```

## compare randomized subsets


```{r compareRepNr, fig.height=10, fig.width=5, dev='svg', echo=F}
withinRnksUnnested %>% 
  filter( !is.na(rnks) ) %>%  # remove NA values
  mutate( logRnks = -log10(1.0001-rnks)) %>% 
  ggplot( aes( x=repNr, y=logRnks) ) +
  facet_wrap(~as.factor(spc1), ncol=1) +
  geom_violin() +
  stat_summary(fun.y=median, geom="point", size=2, shape=23)

```


## compare ref.orthos

```{r compRefOrthos, fig.height=4, fig.width=9, dev='svg', echo=F}
withinRnksUnnested2 <- 
  withinRnks %>%
  mutate(filename=NULL) %>% 
  mutate(rnks = rnks %>% 
           map(map,unlist) %>%  # unlist inner list (transposed ranks)
           map(as_data_frame) %>% 
           map(gather, key="refOrthoSpc",value="rnks")
         ) %>% 
  unnest()

withinRnksUnnested2 %>% 
  filter( !is.na(rnks) ) %>%  # remove NA values
  mutate( logRnks = -log10(1.0001-rnks)) %>% 
  ggplot( aes( x=refOrthoSpc, y=logRnks) ) +
  facet_wrap(~as.factor(spc), nrow=1) +
  geom_violin() +
  stat_summary(fun.y=median, geom="point", size=2, shape=23)
```



```{r nRefcompare, dev='svg', echo=F}
refOrthos <- 
  dir("data/subsets/withinSpecies", recursive = T,
    pattern="refOrthos.RDS",full.names = T) %>% 
  set_names(sub(".*/(..)/refOrthos.RDS","\\1",.)) %>% 
  map( readRDS ) # read refOrthos geneIDs

nRefOrthos <- 
  refOrthos %>% 
  map( map, length ) %>% # count number of refOrthos
  map_df(as_data_frame,.id = "spc") %>% # convert to data_frame
  gather( -spc, key="refOrthoSpc", value="n", na.rm = T)

withinRnksUnnested2 %>% 
  group_by(spc,refOrthoSpc) %>% 
  filter( !is.na(rnks) ) %>%  # remove NA values
  summarise( medLogRnk = median(-log10(1.0001-rnks))) %>% 
  right_join(nRefOrthos, by = c("spc", "refOrthoSpc")) %>% 
  ggplot( aes(x=n,y=medLogRnk)) +
  geom_point( aes(shape = factor(refOrthoSpc), colour=factor(spc)) )
  
```


## rank stability

Do the same genes tend to have same ranks regardless of choice of sample subset? The plot below shows the correlation between the ranks in the subset replicates. There is a clear tendency to correlate even though it varies between the replicates, so in general there seems to be certain genes that tend to have higher ranks than others. This is expected as there are several gene specific properties that can affects its likelihood to get a high rank. Example of properties that are necessary to achieve a stable high rank:

* Many co-expressed genes
* Expressed in many conditions
* Large range of expression across samples.

There seems to be very little effect of ref.orthos and while there is differences between the species, it is far less than the variation between replicates.

```{r rankCorr, fig.height=4, fig.width=9, dev='svg', echo=F}

# helper function to calculate the correlation between all the ranks
rnksCor <- function(rnksList){
  # Because of missing expression values in some subsets there can be different number of genes 
  # in the ORS vectors. 
  
  # Turn list of named ORS vectors into matrix and correlate
  map( rnksList, t) %>%           # transpose each vector to form a row matrix
    rbind.fill.matrix %>%   # bind the row matrices by names and fill missing values with NA
    t %>% 
    cor(use = 'pairwise.complete.obs') %>%  # correlate ranks
    .[lower.tri(.)]   # keep only lower triangle
}


withinRnks %>% 
  unnest1( .id = "refOrthoSpc" ) %>% 
  unnest1( .id = "T" ) %>% 
  group_by( spc, refOrthoSpc, T) %>%  # group together all reps per spc/refOrthoSpc/T
  nest(rnks,.key = rnkCor) %>%    # nest the reps together in a list
  mutate(rnkCor = map(rnkCor, ~ rnksCor(.x$rnks))) %>%
  unnest() %>% 
  ggplot( aes( x=refOrthoSpc, y=rnkCor)) +
  geom_boxplot() +
  facet_wrap(~as.factor(spc), nrow=1)

```


